PERs from Partial Projections for Higher-order Polymorphic Demand Analysis

### 2017-04-17 pubs/demand_analysis/new.tex

%% Kluwer Small Journal Article Sample File
%% June 20, 1994

%%   Use one of these two commands: 
\documentstyle{smjrnl} % For Computer Modern Fonts, or
%\documentstyle[smjfonts]{smjrnl} % for PostScript fonts

\begin{document}
\begin{article}

%%%%% To be entered at Kluwers: =====>>

\journame{Lisp and Symbolic Computation}
\volnumber{?}
\issuenumber{?}
\issuemonth{?}
\volyear{1995}

%% Do not delete either of the following two commands.
%% Please supply facing curly brackets for the part you
%% are not using for this article.
%\received{May 1, 1991}\revised{}

\authorrunninghead{K. Davis}
\titlerunninghead{PERs from Projections for Binding-time Analysis}

@@

%% <<== End of commands to be entered at Kluwers

%%  Authors, start here ==>>

@@

\com{\tmn}{\theta_{mn}}
\com{\tnm}{\theta_{nm}}
\com{\tmy}{\theta_{m\infty}}
\com{\tyy}{\theta_{\infty\infty}}
\com{\tiy}{\theta_{i\infty}}
\com{\tym}{\theta_{\infty{}m}}
\com{\tyz}{\theta_{\infty{\rm0}}}
\com{\tyo}{\theta_{\infty{\rm1}}}
\com{\tyi}{\theta_{\infty{}i}}
\com{\tyip}{\theta_{\infty{}i+{\rm1}}}
\com{\th}{\theta}

@@

\title{PERs from Projections for Binding-time Analysis}

\authors{Kei Davis}
\email{kei@crl.nmsu.edu}

\affil{Center for Non-Linear Studies \\
Mail Stop B258 \\
Los Alamos National Laboratory \\
Los Alamos, NM  87545, U.S.A.}

%\editor{Dennis Kibler}

\abstract{{\it First-order\/} projection-based binding-time analysis
has proven genuinely useful in partial evaluation
\cite{Lau91a}.  There have been three notable generalisations
of projection-based analysis to higher order.  The first lacked a
formal basis \cite{Mog89}; the second used structures strictly more
general than projections, namely {\it partial equivalence relations\/}
(PERs) \cite{HS91}; the third involved a complex construction that
gave rise to impractically large abstract domains \cite{Dav93b}.  This
paper presents a technique free of these shortcomings: it is simple,
entirely projection-based, satisfies a formal correctness condition,
and gives rise to reasonably small abstract domains.  Though the
technique is cast in terms of projections, there is also an
interpretation in terms of PERs.  The principal limitation of the
technique is the restriction to {\it monomorphic\/} typing.}

\keywords{Binding-time analysis, abstract interpretation, projections,
PERs.}

@@

We take as given that binding-time analysis is essential for good
partial evaluation, and we do not address the issue of annotating
programs according to the results of analysis.  Numerous binding-time
analysis techniques have been proposed and implemented; we greatly
narrow the field of discussion by restricting attention to those for
which there is a formally stated notion of correctness that the
technique has been proven to satisfy.

@@

\ \\ \noindent {\bf The Central Result.} For expression \mbox{\tt e} of type
\mbox{\tt T} and free-variable environment type \mbox{\tt E}, the functions
$\Ee{\Ps}{\mbox{\tt e}}$, $\Ee{\Ss}{\mbox{\tt e}}$, and $\Ee{\Ss}{\mbox{\tt e}}$ are logically
related by $\cR{}$, that is
\beqs
\it (\Re{}{\makebox{\tt E}}\fto\Re{}{\makebox{\tt T}})\ (\Ee{\Ps}{\makebox{\tt e}},\ \Ee{\Ss}{\makebox{\tt e}},\ \Ee{\Ss}{\makebox{\tt e}})\ .
\eeqs
Further, $\Re{}{\mbox{\tt T}}~((\A,\K),\cdot,\cdot)$ is a PER for all $\A$ and
$\K$.\ \ $\Box$

@@

We have successfully generalised Launchbury's monomorphic
projection-based binding-time analysis to higher-order, using abstract
domains smaller than those typically used in BHA strictness analysis.

The next step would be to generalise to handle Hindley-Milner
polymorphism.  This has been done with good results at first order for
binding-time analysis \cite{Lau91a}, and for BHA strictness analysis
at higher order \cite{Bar91,Bar93}.  We anticipate that the combined
use of these theories will give a reasonably straightforward
generalisation to polymorphism.  In addition to making the analysis
more widely applicable, this should also greatly reduce the run-time
cost of analysis, as it does for BHA analysis.

Though not developed here, using the same approach it is possible to
give a strictness analysis technique, again closely related to Hunt's
PER-based strictness analysis technique \cite{Hun91}.

@@

\end{article}
\end{document}

>>>>

\documentclass[11pt]{article}

\usepackage{geometry}
\geometry{textwidth=7.5in,
  textheight=9in}

%\usepackage{draftwatermark}
%\SetWatermarkText{DRAFT}
%\SetWatermarkScale{6}
%\SetWatermarkColor{yellow}

\begin{document}

\title{PERs from Partial Projections for Higher-order Polymorphic Demand Analysis}
\author{Kei Davis\\
  Los Alamos National Laboratory\\
  kei at lanl.gov}
\maketitle

@@

@@

\com{\tmn}{\theta_{mn}}
\com{\tnm}{\theta_{nm}}
\com{\tmy}{\theta_{m\infty}}
\com{\tyy}{\theta_{\infty\infty}}
\com{\tiy}{\theta_{i\infty}}
\com{\tym}{\theta_{\infty{}m}}
\com{\tyz}{\theta_{\infty{\rm0}}}
\com{\tyo}{\theta_{\infty{\rm1}}}
\com{\tyi}{\theta_{\infty{}i}}
\com{\tyip}{\theta_{\infty{}i+{\rm1}}}
%\com{\th}{\theta}

@@

\begin{abstract}
  {\it First-order\/} projection-based strictness analysis (properly,
  \emph{demand analysis}) was first introduced by Wadler and Hughes in
  1987~\cite{WH87}.  This development was restricted to first-order,
  monomorphic languages and provided no proof of soundness.  Subsequent
  developments extended the theory to other analyses, approaches to higher
  order, and polymorphism.  Yet other developments were based on
  generalization of projections, specifically partial equivalence relations
  and partial projections, the latter notably simplifying a difficulty with
  projection-based strictness dating back to Wadler and Hughes.  However, to
  date no development has combined the desirable contributions of all of these
  approaches.  We present the first practical higher-order, polymorphic,
  partial-projection/partial-equivalence-relation based demand analysis,
  together with a proof of satisfaction of the fundamental safety condition
  for such analyses.  Such an analysis can be used both to improve serial code
  generation, as well as enable safe automatic parallelization, for lazy or
  non-strict functional languages.

% has proven genuinely useful in partial evaluation
% \cite{Lau91a}.  There have been three notable generalisations
% of projection-based analysis to higher order.  The first lacked a
% formal basis \cite{Mog89}; the second used structures strictly more
% general than projections, namely {\it partial equivalence relations\/}
% (PERs) \cite{HS91}; the third involved a complex construction that
% gave rise to impractically large abstract domains \cite{Dav93b}.  This
% paper presents a technique free of these shortcomings: it is simple,
% entirely projection-based, satisfies a formal correctness condition,
% and gives rise to reasonably small abstract domains.  Though the
% technique is cast in terms of projections, there is also an
% interpretation in terms of PERs.  The principal limitation of the
% technique is the restriction to {\it monomorphic\/} typing.
\end{abstract}

%\keywords{Binding-time analysis, abstract interpretation, projections, PERs.}

@@

We take as given that demand analysis is essential for good serial code
generation, and for identifying where parallelism may be safely introduced.
Numerous such analysis techniques have been proposed and implemented; we
greatly narrow the field of discussion by restricting attention to those for
which there is a formally stated notion of correctness that the technique has
been proven to satisfy.

@@

\ \\ \noindent {\bf The Central Result.} For expression \mbox{\tt e} of type
\mbox{\tt T} and free-variable environment type \mbox{\tt E}, the functions
$\Ee{\Ps}{\mbox{\tt e}}$, $\Ee{\Ss}{\mbox{\tt e}}$, and $\Ee{\Ss}{\mbox{\tt e}}$ are logically
related by $\cR{}$, that is
\beqs
\it (\Re{}{\makebox{\tt E}}\fto\Re{}{\makebox{\tt T}})\ (\Ee{\Ps}{\makebox{\tt e}},\ \Ee{\Ss}{\makebox{\tt e}},\ \Ee{\Ss}{\makebox{\tt e}})\ .
\eeqs
Further, $\Re{}{\mbox{\tt T}}~((\A,\K),\cdot,\cdot)$ is a PER for all $\A$ and
$\K$.\ \ %$\Box$

@@

We have successfully generalised Launchbury's polymorphic
partial-projection-based demand analysis to higher-order, using abstract
domains smaller than those typically used in BHA strictness analysis.

%The next step would be to generalise to handle Hindley-Milner
%polymorphism.  This has been done with good results at first order for
%binding-time analysis \cite{Lau91a}, and for BHA strictness analysis
%at higher order \cite{Bar91,Bar93}.  We anticipate that the combined
%use of these theories will give a reasonably straightforward
%generalisation to polymorphism.  In addition to making the analysis
%more widely applicable, this should also greatly reduce the run-time
%cost of analysis, as it does for BHA analysis.

%Though not developed here, using the same approach it is possible to
%give a strictness analysis technique, again closely related to Hunt's
%PER-based strictness analysis technique \cite{Hun91}.

@@

\end{document}
